---
title: "Maternal Health Risk Report"
author: "Omar Abdul-Rahman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
# Ensure required packages are installed
packages <- c("tidyverse", "gtsummary", "ggplot2", "janitor", "here", "broom.helpers")
new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new_packages)) install.packages(new_packages, dependencies = TRUE)

library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(gtsummary)
library(ggplot2)
library(janitor)
library(here)
library(broom.helpers)

# Loading dataset

data <- read_csv("data/maternal_riskdata.csv")

# Checking data

data <- janitor::clean_names(data)
head(data)
cat("Rows:", nrow(data), "Cols:", ncol(data), "\n")
head(data)
```

This dataset came from the Maternal Health Risk Data Set. This contains clinical measurements collected from pregnant women.

It includes variables such as age, systolic and diastolic blood pressure, blood sugar, body temperature, and heart rate, along with an overall risk level indicator (low, mid, or high).

The goal of this analysis is to identify which physiological factors are most associated with maternalhealth data

```{r analysis}
# Converts risk level to factor

data <- data %>%
  mutate(
    risk_level = str_to_lower(str_replace_all(risk_level, " ", "")),
    risk_level = factor(risk_level, levels = c("lowrisk", "midrisk", "highrisk"))
  )


# The summary table by risk level

tbl_summary(
  data = data,
  by = risk_level,
  label = list(
    age ~ "Age (years)",
    systolic_bp ~ "Systolic BP",
    diastolic_bp ~ "Diastolic BP",
    bs ~ "Blood Sugar",
    body_temp ~ "Body Temp (°C)",
    heart_rate ~ "Heart Rate"
  ),
  statistic = all_continuous() ~ "{mean} +/- {sd}"
) %>%
  add_overall() %>%
  modify_header(label ~ "**Variable**") %>%
  modify_caption("**Table 1.** Summary of maternal health indicators by risk level")
 
# My plot

ggplot(data, aes(x = risk_level, y = systolic_bp, fill = risk_level)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Systolic Blood Pressure by Risk Level",
    x = "Maternal Risk Level",
    y = "Systolic BP (mm Hg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
  
# This predicts high risk vs low/mid

data <- data %>%
  mutate(high_risk = ifelse(risk_level == "high", 1, 0))

# This is the fit logistic regression model

model <- glm(
  high_risk ~ age + systolic_bp + diastolic_bp + bs + body_temp + heart_rate,
  data = data,
  family = binomial()
)

# This display model results (odds ratios)
tbl_regression(model, exponentiate = TRUE) %>%
  modify_caption("**Table 2.** Logistic regression predicting high maternal health risk")

cat("
- Higher systolic and diastolic blood pressure were associated with higher maternal risk.
- Elevated BMI and blood sugar also appear linked to risk classification.
- These findings highlight the importance of routine monitoring of blood pressure and glucose in maternal care.
")
```